---
title: "Homework2"
author: "Yunjia Liu"
date: "2024-09-30"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

## Problem1

### 1. Process the data
Input data and do using clean_names() to clean the data. Using select() to choose the required columns. Specially when treating all the route1~route11, I use the starts_with() function to simplify the process. And then I use mutate to convert 'entry' from character to logical variable.
```{r}
nyctransit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude, starts_with("route"), entry, vending, entrance_type, ada) |>
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

head(nyctransit_df)

num_rows = nrow(nyctransit_df)
num_cols = ncol(nyctransit_df)
cat("The dataset has", num_rows, "rows and", num_cols, "columns.\n")
```

### 2. Summary:
2.1. Variables in the data:
  - line: The subway line the station belongs to.
  - station_name: Name of the station.
  - station_latitude / station_longitude: Geographic coordinates of the station.
  - routes1~routes11: Subway routes served by the station.
  - entry: whether the entrance allows entry.
  - vending: Whether the entrance has a vending machine.
  - entrance_type: Type of entrance 
  - ada: whether the entrance is ADA compliant.

2.2. Cleaning steps:
After using clean_names() to rename the variables, I retained the relevant columns, converted the entry column to a logical variable, and inspected the data for consistency. The dataset contains 1868 rows and 19 columns. As far as I am concerned, the data is relatively neat because each row corresponds to an entry and each variable has a different value.


### 3. Quesiton
3.1 Number of distinct stations.
```{r}
distinct_stations = nyctransit_df |>
  distinct(line, station_name)

num_distinct_stations <- nrow(distinct_stations)
cat("Number of distinct stations: ", num_distinct_stations, "\n")
```

3.2 Number of stations that are ADA compliant.
```{r}
ada_compliant = nyctransit_df |>
  filter(ada == "TRUE") |>
  distinct(line, station_name)

num_ada_compliant = nrow(ada_compliant)
cat("Number of ADA compliant stations: ", num_ada_compliant, "\n")
```

3.3 Proportion of entrances without vending allow entrance.
```{r}
no_vending_entry = nyctransit_df |>
  filter(vending == "NO") |>
  summarise(proportion_entry = mean(entry))

cat("Proportion of entrances without vending allow entrance: ", no_vending_entry$proportion_entry, "\n")
```

3.4 Question4

Reformat the dataset. 
Using mutate(across(...)) to convert all route columns to characters so that all route1~route11 columns are characters.
Using pivot_longer to combine all route1~route11 columns into a single "route_name" column and make the original column as "route_number".(I am bit confuse about the description)
```{r}
# Use pivot_longer to combine all route1~route11 columns into a single "route" column

transit_reformat = nyctransit_df |>
  mutate(across(starts_with("route"), as.character)) |>
   pivot_longer(route1:route11, names_to = "route_number", values_to = "route_name", values_drop_na = TRUE)
head(transit_reformat)
```

Filter the dataset to find stations that serve the A train and count the number of distinct stations serving the A train
```{r}
a_train_stations = transit_reformat |>
  filter(route_name == "A") |>
  distinct(line, station_name)

num_a_train_stations = nrow(a_train_stations)
cat("Number of distinct stations serving the A train: ", num_a_train_stations, "\n")
```

Find the number of A train stations that are ADA compliant and count the number of ADA compliant stations serving the A train.
```{r}
ada_compliant_a_train = transit_reformat |>
  filter(route_name == "A") |>
  filter(ada == "TRUE") |>
  distinct(line, station_name)

num_ada_compliant_a_train = nrow(ada_compliant_a_train)
cat("Number of ADA compliant stations serving the A train: ", num_ada_compliant_a_train, "\n")
```


## Problem2

### 1. Process the data
1.1 Input the data in 'Mr. Trash Wheel' Sheet.
```{r}
mr_trash_wheel =
  read_excel("./data/202409_Trash_Wheel_Collection_Data.xlsx", sheet = "Mr. Trash Wheel", range = cell_cols("A:N"), skip = 1, na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  drop_na(dumpster)|>
  mutate(
    sports_balls = as.integer(round(sports_balls, 0)),  # Round and convert to integer
    trash_wheel = "Mr. Trash Wheel",
    year = as.integer(year))
```


1.2 Read and clean Professor Trash Wheel sheet & Gwynnda's data sheet. Combine all three seperate datasets together.
```{r}
prof_trash_wheel = read_excel("./data/202409_Trash_Wheel_Collection_Data.xlsx", 
                                sheet = "Professor Trash Wheel", 
                                skip = 1,  # Adjust based on where data starts
                                col_names = TRUE,
                                na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  drop_na(dumpster)|>
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda_trash_wheel = read_excel("./data/202409_Trash_Wheel_Collection_Data.xlsx", 
                                   sheet = "Gwynnda Trash Wheel", 
                                   skip = 1,
                                   col_names = TRUE,
                                  na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  drop_na(dumpster)|>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda Trash Wheel"
  )

# Combine datasets into one tidy dataset
combined_trash_data <- bind_rows(mr_trash_wheel, prof_trash_wheel, gwynnda_trash_wheel)
```


### 2. Answer to the question: 
2.1.total weight of trash collected by Professor Trash Wheel
```{r}
total_weight_prof = combined_trash_data |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarise(total_weight = sum(weight_tons, na.rm = TRUE))
```

2.2. total number of cigarette butts collected by Gwynnda in June 2022:
```{r}
total_cig_butts_gwynnda = combined_trash_data |>
  filter(trash_wheel == "Gwynnda Trash Wheel" & month(date) == 6 & year == 2022) |>
  summarise(total_cig_butts = sum(`cigarette_butts`, na.rm = TRUE))

```

### 3. Summary:
```{r}
n_observations <- nrow(combined_trash_data)
total_weight_prof <- total_weight_prof$total_weight
total_cig_butts_gwynnda <- total_cig_butts_gwynnda$total_cig_butts

cat("The combined dataset includes", n_observations, "observations from three Trash Wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. 
Key variables in the dataset include `weight_tons` (the weight of trash collected in tons) and `cigarette_butts`. 
For example, Professor Trash Wheel has collected a total of", total_weight_prof, "tons of trash. 
In June 2022, Gwynnda collected a total of", total_cig_butts_gwynnda, "cigarette butts.")
```


## Problem3

